Governing through patronage: the bargain for education in decentralized Brazil

### 2020-09-10 paper/paper.tex
add additional plots

\documentclass[12pt,a4paper]{article}
\usepackage{setspace}
\usepackage[top = 1.25in, bottom = 1.25in, left = 1.25in, right = 1.25in]{geometry}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{graphicx}
\usepackage{natbib}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{float}
\usepackage{pdflscape}
\usepackage{booktabs}
\usepackage{dcolumn}
\usepackage{pdflscape}
\usepackage[hyphens]{url}
\usepackage{enumitem}
\usepackage[table]{xcolor}
\usepackage{authblk}
\usepackage{appendix}
\usepackage{titletoc}
\setlength\parindent{20pt}
\usepackage{setspace}
\usepackage{multirow}
\usepackage{caption}
\usepackage{titling}
\usepackage{amsmath}
\usepackage{hyperref}

@@

\bibliographystyle{unsrtnat}
\begin{document}

\section{Introduction}

@@

The quality of public education in the across the world, and particularly in the developing world, remains low: functional illiteracy is widespread, and children do not master basic arithmetic operations. How can these educational outcomes be improved? Previous studies have found that incentives, whether these be electoral or performance-related, can help raise the quality of public services \citep{gulzar_politicians_2017,duflo_incentives_2012}. Often relying on quasi-experimental research designs, these studies provide compelling evidence that actors respond to policy interventions.\citep{finan_personnel_2015} But it remains unclear how these incentives connect to the broader institutional environment: who delivers public services and what political institutions shape their behavior?

@@

I first turn to testing the main proposition of this paper: that turnover in educational staff stems indeed from coalitional dynamics. There is strong evidence that mayors who control more (less) seats in the city council resort to less (more) patronage. Figure \ref{tbl:turnover} presents the results of the estimation. The first model regresses turnover index on coalition share, where the unit of analysis is a grade-level per school. Models 2 and 3 estimate the effect of the share of legislative seats on hiring or dismissal of individual bureaucrats. The third set of models (4 and 5) estimate the change in the share of new hires and dismissals at the municipal level. All models suggest that an increase in the executive share of legislative seats decrease turnover.

@@

% For visual intuition, I present the predicted values for models 4 and 5, respectively the share of hired and dismissed bureaucrats against the share of executive seats in the city council. Albeit by construction the bivariate relationship is linear, it demonstrates that the amount of patronage is decreasing as executive coalition controls more seats in the legislature. Overall, these results suggest that weaker executive control over the legislature increases patronage, with potentially negative effects for public service delivery in municipalities in Brazil.

% \begin{figure}[H]
%     \centering
%     \includegraphics{plots/plot_pred}
%     \caption{Predicted values for bureaucratic turnover. The graph plots the proportion of educational staff either hired or fired for any given year against the share of seats controlled by the executive. The circles' size is proportional to the number of observations contained at each value of the legislative share of seats.}
%     \label{fig:turnover}
% \end{figure}

>>>>

\documentclass[12pt,a4paper]{article}
\usepackage{setspace}
\usepackage[top = 1in, bottom = 1in, left = 1in, right = 1in]{geometry}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{graphicx}
\usepackage{natbib}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{float}
\usepackage{pdflscape}
\usepackage{booktabs}
\usepackage{dcolumn}
\usepackage{pdflscape}
\usepackage[hyphens]{url}
\usepackage{enumitem}
\usepackage[table]{xcolor}
\usepackage{authblk}
\usepackage{appendix}
\usepackage{titletoc}
\setlength\parindent{20pt}
\usepackage{setspace}
\usepackage{multirow}
\usepackage{caption}
\usepackage{titling}
\usepackage{amsmath}
\usepackage{hyperref}

@@

\bibliographystyle{unsrtnat}

\title{Governing through patronage: the bargain for education in decentralized Brazil}
\date{\today}

\author{Galileu Kim}
\affil{Princeton University}

\begin{document}

\maketitle

\begin{abstract}
    Why do politicians engage in patronage and what are its consequences for public service delivery? I find that staff turnover in educational staff results from mayors attempt to exchange public sector jobs for support from local legislators. Leveraging fieldwork evidence and administrative data, I demonstrate that mayors with a weaker hold on the local legislature resort to patronage to co-opt legislators. Educational staff turnover has negative downstream effects on student learning, but limited electoral backlash. Weak electoral accountability channels suggests that patronage is primarily a political elite game, with little involvement by the electorate. These findings point to the dangers of elite capture of public services and its downstream consequences for social welfare.
\end{abstract}

\newpage
\section{Introduction}

@@

The quality of public education in the across the world, and particularly in the developing world, remains low: functional illiteracy is widespread, and children do not master basic arithmetic operations. How can these educational outcomes be improved? Previous studies have found that individual incentives, whether these be electoral or performance-related, can help raise the quality of public services \citep{gulzar_politicians_2017,duflo_incentives_2012}. Often relying on quasi-experimental research designs, these studies provide compelling evidence that political and bureaucratic actors respond to policy interventions.\citep{finan_personnel_2015} But it remains unclear how these incentives connect to the broader institutional environment: who delivers public services and what political institutions shape their behavior?

@@

I first turn to testing the main proposition of this paper: that turnover in educational responds to the degree of political alignment between the mayor and politician. There is strong evidence that mayors who control more (less) seats in the city council resort to less (more) patronage. Figure \ref{tbl:turnover} presents the results of the estimation. The first model regresses turnover index on coalition share, where the unit of analysis is a grade-level per school. Models 2 and 3 estimate the effect of the share of legislative seats on hiring or dismissal of individual bureaucrats. The third set of models (4 and 5) estimate the change in the share of new hires and dismissals at the municipal level. All models suggest that an increase in the executive share of legislative seats decrease turnover.

@@

For visual intuition, I present the predicted values for models 4 and 5, respectively the share of hired and dismissed bureaucrats against the share of executive seats in the city council. Albeit by construction the bivariate relationship is linear, it demonstrates that the amount of patronage is decreasing as executive coalition controls more seats in the legislature. Overall, these results suggest that weaker executive control over the legislature increases patronage, with potentially negative effects for public service delivery in municipalities in Brazil.

\begin{figure}[h]
    \centering
    \includegraphics{plots/plot_interactions_logit}
    \caption{Predicted values for bureaucratic turnover. The graph plots the proportion of educational staff either hired or fired for any given year against the share of seats controlled by the executive. The circles' size is proportional to the number of observations contained at each value of the legislative share of seats.}
    \label{fig:turnover}
\end{figure}
